---
title: "PubScore"
author: "Tiago Lubiana"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: readable
    highlight: tango
    fig_width: 17
    fig_height: 10
    toc: false
vignette: >
  %\VignetteIndexEntry{FCBF :  Fast Correlation Based Filter for Feature Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = TRUE}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

PubScore is an R package with 2 goals :
- (1) provide a quantitative score of the relevance of a list of genes regarding any topic a
- (2)  visualize combinations of genes and terms of interest interactively.

Querying the PubMed, PubScore gets the article count for  gene/term combination and creates a  score to calculate the literature enrichment for a list of genes.


Let's start by installing the package. 


Let's explore how to calculate a literature score for selected genes, 
and estimate the likelihood of such a score (or higher) happening by chance. 

We will use the scDengue dataset, of dengue-infected monocytes, from the FCBF package. 
We will see the FCBF selected genes are already related in the literature to dengue. 

```{r}
BiocManager::install("PubScore")

library("FCBF")
library('SummarizedExperiment')
library(PubScore)
# load expression data
data(scDengue)
```

Let's run the FCBF algorithm to select genes that might be relevant 
to understand


```{r}
infection <- SummarizedExperiment::colData(scDengue)
target <- infection$infection

exprs <- as.data.frame(SummarizedExperiment::assay(scDengue,'logcounts'))
discrete_expression <- as.data.frame(FCBF::discretize_exprs(exprs))

fcbf_features <- fcbf(discrete_expression, target, thresh = 0.05, verbose = FALSE)
total_features <- rownames(exprs)
head(fcbf_features)
```



The scDengue dataset is using Ensemble IDs, and usually papers use gene symbols 
(as TP53 and CD3D). 

This is important, as PubScore will search for the genes on pubmed EXACTLY as they are on your vector. The package biomaRt can be used to fix this, as it contains the reference databases we need. 

```{r}
library(biomaRt)

ensembl = useMart("ensembl",dataset="hsapiens_gene_ensembl")
reference = biomaRt::getBM(attributes = c('ensembl_gene_id', 'hgnc_symbol'), 
              mart = ensembl)

total_genes <- reference[reference$ensembl_gene_id %in% total_features,]$hgnc_symbol

fcbf_genes<- reference[reference$ensembl_gene_id %in% rownames(fcbf_features),]$hgnc_symbol
```

Unfortunately, not all genes in the datasets had an equivalent gene symbol.
PubScore does not deal with empty gene-names, so we need to remove them

```{r}
total_genes <- total_genes[total_genes != ""]
fcbf_genes <- fcbf_genes[fcbf_genes != ""]
```


First let's visualize the association of the fcbf_genes with dengue-related terms.

The PubScore package queries the PubMed database, so it takes about 0.7s for 
each search. 

As we have 62 genes and 2 terms, 62 * 0.7 = 43.8 s, the function should
take less than a  minute to run.

After getting the literature object from the get_literature score function,  we can visualize the gene-term relationships with the plot_literature_score function. 


```{r}
terms_of_interest <- c('Dengue')

literature_object <- PubScore::get_literature_score(gene = fcbf_genes,is.list = T,terms_of_interest = terms_of_interest)
library(ggplot2)

p <- plot_literature_score(plot_counts = literature_object$counts, return_ggplot = T)
library(plotly)
ggplotly(p)
```


There are a few genes that have been related to dengue previously, such as MAVS. There is a Nature Immunology research highligh entitled "Dengue suppresses MAVS". 
(www.nature.com/articles/ni.3527)

FES may be a false positive, but MX1 is related to antiviral response and NFKB2 is related to inflamation.


Qualitatetively, it seems that indeed some of these genes are already related to Dengue. 
But how many Dengue-related genes would I expect by chance alone?

For this, we need to estimate the null distribution of the scores by randomly selecting
genes. That is the goal of the 'test_literature_score' function. 

WARNING: This functions takes 6 hours to run. That is because a lot of requests are made to
pubmed. We ran it and made the final output available here so you don't need it. 
It is commented out of the code because of that, but feel free to run it. 

```{r eval = FALSE} 

#literature_object <- test_literature_score(literature_object, total_genes, remove_ambiguous = F )
#save(literature_object, file= '../literature_object.RData')
load('../literature_object.RData')
```


We will re-do the p-value estimation just to be really clear about how it is done. 
```{r}

# A table with all possible combinations is generated
simulation_of_literature_null <- literature_object$all_gene_combinations

# Then, a number of genes equal to your original list is picked
# and, from this list, we get a score.
# This process is repeated 10 thousand times.

message('Running 10000 simulations')
      distribution_of_scores <- c()
      for (i in seq_len(10000)){
        
        genes_to_sample_now <- sample(total_genes, 62)
        
        simu_now <- simulation_of_literature_null[simulation_of_literature_null$Gene %in% genes_to_sample_now,]$count
        
        list_score <- sum(simu_now) / 62
        distribution_of_scores <- c(distribution_of_scores,list_score )
        
      }
      
    distribution_of_scores<- data.frame(d = distribution_of_scores)
    
    # The score to compare is the one of the original list  
    score <- literature_object$list_literature_score
      
    # Then you ask: how many of the null simulations had a score
    # as high as the one of the original list?
    # This is the simulated p-value!

    pvalue <-sum(distribution_of_scores[,1] >= score)/length(distribution_of_scores[,1])
    
    
    print('The p-value by simulation is:')
    print(pvalue) 
      
```

Hm, the pvalue indicates that the list of genes obtained is not more related 
to Dengue than what we would expect by chance. This might be a sign that FCBF
did not work very well for this discretization/dataset. Or that the information 
regarding the association of these genes to Dengue  is largely  absent from the 
literature.

Let's look at the top genes associated with Dengue in the literature, in the literature object.


```{r}
literature_object$all_gene_combinations$Gene[order(literature_object$all_gene_combinations$count, decreasing = T)][1:30]
```


Hm there are clearly problems with these genes... they are ambiguous!
We will have to blacklist a few genes to avoid these outliers. 
After some digging, I manually selected this list of ambiguously named genes.

Genes blacklisted :      
"PC", "JUN", "IMPACT", "ACHE", "SRI", "SET", "CS", "PROC", 
"MET", "SHE", "CAD", "DDT", "PIGS", "SARS", "REST", "GC", "CP", 
"STAR", "SI", "GAN", "MARS", "SDS", "AGA", "NHS", "CPE", "POR", 
"MAX", "CAT", "LUM", "ANG", "POLE", "CLOCK", "TANK", "ITCH", 
"SDS", "AES", "CIC", "FST", "CAPS", "COPE", "F2", "AFM", "SPR", 
"PALM", "C2", "BAD", "GPI", "CA2", "SMS", "INVS", "WARS", "HP", 
"GAL", "SON", "AFM", "BORA", "MBP", "MAK", "MALL", "COIL", "CAST"" 


Instead of running the time-consuming test_literature_score function again,
we can use the much quicker "retest_literature_score" function.

```{r}

retest_literature_score(literature_object, remove_ambiguous = T)
```

0.04266 Ok, it is a higher score than expected by random in 95.7% of the cases. 
The threshold of 95% is rather arbitrary and there is a big movement to reject the hard definition of statistical significance.

In this case, the p-value is shorter than the 0.05, so I can say confidently that the list returned by FCBF is more related to Dengue than expected by chance. 



